---
title: "Lab 6"
author: "PSTAT126"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library("alr4")
library("MASS")
```

### Consider the variable selection for a regression model:

From Highway data in package we try to model the 'rate' variable (accident rates per million vehicle miles, 1973) on other predictors.

After checking for correlaton of the response rate with the other numerical variables, the chosen top 3 covariates for the model are: acpt (0.752), sigs (0.564), slim (-0.681), the correlation with the response are given in the parentheses. 

```{r}
pairs(Highway[,c("acpt","sigs","slim","rate")])
fit<- lm(rate~ acpt + sigs + slim + acpt:sigs + acpt:slim + sigs:slim , data = Highway) # We include all the interactions

summary(fit)
```
Remove the predictor with highest p-value iteratively;

```{r}
fit<- lm(rate~ acpt + sigs + slim + acpt:sigs + sigs:slim , data = Highway) # We remove acpt:slim 
summary(fit)

fit<- lm(rate~ acpt + sigs + slim + acpt:sigs , data = Highway)  
summary(fit)

fit<- lm(rate~ acpt + sigs + slim, data = Highway)
summary(fit)

fit<- lm(rate~ acpt + slim, data = Highway)
summary(fit)

```

## Diagnostic plots:

```{r}
par(mfrow=c(2,2))
plot(fit$fitted.values, fit$residuals)
abline(h=0, col="red")

plot(Highway$acpt, fit$residuals)
abline(h=0, col="red")

plot(Highway$slim, fit$residuals)
abline(h=0, col="red")

qqnorm(fit$residuals)
qqline(fit$residuals, col="red")

```

```{r}
par(mfrow=c(2,2))
plot(hatvalues(fit), cooks.distance(fit), xlab="Leverages", ylab="Cook's Distance")

plot(hatvalues(fit), type = 'h',xlab = "Observation")

plot(cooks.distance(fit), type = "h", xlab ="Observation")

```

## Box-Cox transformation
```{r}
boxcox(fit, plotit=T, lambda=seq(-1,1,len=100))
```

Therefore, take log transformation on the response

```{r}
fit<- lm(log(rate)~ acpt + slim, data = Highway)
summary(fit)

par(mfrow=c(2,2))
plot(fit$fitted.values, fit$residuals)
abline(h=0, col="red")

plot(Highway$acpt, fit$residuals)
abline(h=0, col="red")

plot(Highway$slim, fit$residuals)
abline(h=0, col="red")

qqnorm(fit$residuals)
qqline(fit$residuals, col="red")

```


```{r}
par(mfrow=c(2,2))
plot(hatvalues(fit), cooks.distance(fit), xlab="Leverages", ylab="Cook's Distance")

plot(hatvalues(fit), type = 'h',xlab = "Observation")

plot(cooks.distance(fit), type = "h", xlab ="Observation")

```

